---
title: "Data 607 Assignment 4"
author: "Jayden Jiang"
date: "2025-02-20"
output: html_document
---

## Library
```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
```

## Read the information for .CSV file
```{r}
flights_data <- read.csv ("https://raw.githubusercontent.com/JaydeeJan/Data-607-Assignment-4/refs/heads/main/flights_data.csv")
flights_data
```
## Clean and organize the data
```{r}
flights_clean <- flights_data[!apply(flights_data, 1, function(x) all(is.na(x) | x == "")),]
flights_clean

flights_clean$X[2] <- "ALASKA"
flights_clean$X[4] <- "AM West"
flights_clean

colnames(flights_clean)[1] <- "Airline"
colnames(flights_clean)[2] <- "Status"
colnames(flights_clean)

flights_clean
  
```


## Tidy and transform the data from wide to long format
```{r}

str(flights_clean)

flights_clean$Phoenix <- as.numeric((gsub(",", "", flights_clean$Phoenix)))

flights_clean$Seattle <- as.numeric(gsub(",", "", flights_clean$Seattle))

flights_reshape <- flights_clean %>%
  pivot_longer(cols = -c(Airline, Status),
               names_to = "City",
               values_to = "Flights")
flights_reshape
```

## Filter for delayed flights and compare delayed flights from both airlines
```{r}
delayed_flights <- flights_reshape %>%
  filter(Status == "Delayed")

summary_delays <- delayed_flights %>%
  group_by(City, Airline) %>%
  summarise(Total_Delays = sum(Flights),
            .groups = 'drop')

summary_delays

```
```{r}
ggplot(data = summary_delays, aes(x = City, y = Total_Delays, fill = Airline)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = " Comparison of Delayed Airline by City",
       x = "City",
       y = "Total Delays") +
  theme_minimal()
```


## Filter on-time flights and compare on-time flights from both airlines
```{r}
ontime_flights <- flights_reshape %>%
  filter(Status == "On Time")

summary_ontime <- ontime_flights %>%
  group_by(City, Airline) %>%
  summarise(Total_OnTime = sum(Flights),
            .groups = 'drop')

summary_ontime
```

```{r}
ggplot(data = summary_ontime, aes(x = City, y = Total_OnTime, fill = Airline)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = " Comparison of On-Time Airline by City",
       x = "City",
       y = "Total On-Time") +
  theme_minimal()
```

## Calculate the total and average delays for each airline across all cities.
```{r}
delay_analysis <- delayed_flights %>%
  group_by(Airline) %>%
  summarise(
    Total_Delays = sum(Flights),
    Avg_Delays = mean(Flights)
  )

delay_analysis
```

## Calculate the total and average on-time for each airline across all cities.
```{r}
ontime_analysis <- ontime_flights %>%
  group_by(Airline) %>%
  summarise(
    Total_OnTime = sum(Flights),
    Avg_OnTime = mean(Flights)
  )

ontime_analysis
```

## Data showing total delayed & on-time percentage and average delayed & on-time percentage
```{r}

total_flights <- flights_reshape %>%
  group_by(Airline, City) %>%
  summarize(Total_Flights = sum (Flights),
            .groups = 'drop')

total_flights

flight_percentage <- flights_reshape %>%
  left_join(total_flights, by = c("City", "Airline")) %>%
  mutate(Percentage = (Flights / Total_Flights) * 100)

flight_percentage


# calculating the percentage of delayed and on-time performance for ALASKA Airline

alaska_flights <- flights_reshape %>%
  filter(Airline == "ALASKA")
alaska_flights

alaska_delayed <- alaska_flights %>%
  filter(Status == "Delayed") %>%
  summarize(Delayed_Flights = sum(Flights))
alaska_delayed

alaska_ontime <- alaska_flights %>%
  filter(Status == "On Time") %>%
  summarize(OnTime_Flights = sum(Flights))
alaska_ontime

alaska_total <- alaska_flights %>%
  summarize(Total_Flights = sum(Flights))
alaska_total

alaska_delayed_percentage <- (alaska_delayed$Delayed_Flights / alaska_total$Total_Flights) * 100
alaska_delayed_percentage 

alaska_ontime_percentage <- (alaska_ontime$OnTime_Flights / alaska_total$Total_Flights) * 100
alaska_ontime_percentage 


# Calculating the percentage of delayed and on-time performance for AM West Airline

amwest_flights <- flights_reshape %>%
  filter(Airline == "AM West")
amwest_flights

amwest_delayed <- amwest_flights %>%
  filter(Status == "Delayed") %>%
  summarize(Delayed_Flights = sum(Flights))
amwest_delayed

amwest_ontime <- amwest_flights %>%
  filter(Status == "On Time") %>%
  summarize(OnTime_Flights = sum(Flights))
amwest_ontime

amwest_total <- amwest_flights %>%
  summarize(Total_Flights = sum(Flights))
amwest_total

amwest_delayed_percentage <- (amwest_delayed$Delayed_Flights / amwest_total$Total_Flights) * 100
amwest_delayed_percentage 

amwest_ontime_percentage <- (amwest_ontime$OnTime_Flights / amwest_total$Total_Flights) * 100
amwest_ontime_percentage 


# Comparing both airlines performance

airlines_analysis <- data.frame(
  Airline = c("ALASKA", "AM West"),
  Total_Flights = c(alaska_total$Total_Flights, amwest_total$Total_Flights),
  Delayed_Flights = c(alaska_delayed$Delayed_Flights, amwest_delayed$Delayed_Flights),
  OnTime_Flights = c(alaska_ontime$OnTime_Flights, amwest_ontime$OnTime_Flights),
  Delayed_Percentage = c(alaska_delayed_percentage, amwest_delayed_percentage),
  OnTime_Percentage = c(alaska_ontime_percentage, amwest_ontime_percentage)
)

airlines_analysis

```

## In Conclusion

### Based on the analysis of the flights data for ALASKA and AM West, we can conclude both airlines show a good performance with over 86% of their flights arriving on time. However, AM West Airline appears to have less delays and higher on-time flights. Therefore, AM West offer a slightly more reliable service for passengers who priortized on-time performance. 
